Dispersal and Connectivity in Increasingly Extreme Climatic Conditions

### 2024-07-01 Chapter_2/04_Manuscript/Manuscript.tex
Publish PHD

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
\usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}

%------------------------------------------------------------------------------
%	Some Styling
%------------------------------------------------------------------------------
% Creating some TikZ styles
\tikzset{
  nonterminal/.style = {rectangle
    , minimum size = 6mm
    , very thick
    , draw = black!
  }
}

@@

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{Dispersal Through Seasonal Landscapes}

% \title{Step by Step: The Utility of Dispersal Simulations to Assess Landscape
% Connectivity}

% List of Authors
\author{
  David D. Hofmann\textsuperscript{1,\S} \and
  John W. McNutt\textsuperscript{2} \and
  Arpat Ozgul\textsuperscript{1} \and
  Gabriele Cozzi\textsuperscript{1,2} \and
  Dominik M. Behr\textsuperscript{1,2}
}

@@

% Required to be able to cite
\bibliographystyle{apalike}

@@

\textsuperscript{1} Department of Evolutionary Biology and Environmental
Studies, University of Zurich, Winterthurerstarsse 190, 8057 Zurich,
Switzerland.

\textsuperscript{2} Botswana Predator Conservation, Private Bag 13, Maun,
Botswana.

\textsuperscript{\S} Corresponding author (david.hofmann2@uzh.ch)

@@

\textbf{Running Title:} Release the Dogs! Simulating Wild Dog Dispersal to
Assess Landscape Connectivity

@@

\textbf{Keywords:} dispersal, simulation, movement, integrated step selection,
Kavango-Zambezi Transfrontier Conservation Area, landscape connectivity, Lycaon
pictus

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
This is the abstract.
\end{abstract}

@@

\onehalfspacing
\tableofcontents
\doublespacing

@@

% Taken from Zeller.2020
Connectivity for wildlife is typically modeled in static environments and
estimated for a single snapshot in time (e.g., [6,7]). However, wildlife
responses to landscapes are inherently dynamic, as are the landscapes themselves
(e.g., [8])

Modeling connectivity in a static environment may mask the importance of connectivity for
maintaining wildlife population dynamics and may result in identifying corridors that are only
periodically functional [15,16,18]

\section{Methods}

\section{Results}

>>>>

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{textcomp}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage{hyperref}
\usepackage[nameinlink]{cleveref}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}
\usepackage{scalerel}
\usepackage{makecell}
\usepackage{ifthen}
\usepackage{tikz}
\usepackage{lineno}
\usetikzlibrary{svg.path}

%------------------------------------------------------------------------------
%  Bibtex vs Biblatex
%------------------------------------------------------------------------------
% While bibtex is very nice to use, it's insanely slow. I therefore work with
% bibtex until the final stage. Set the value below to "true" or "false" to pick
% which one should be used.
\newboolean{usebiblatex}
\setboolean{usebiblatex}{true}

% To remove figures from the document, only keeping their captions
\AtBeginDelayedFloats{%
  \renewcommand{\includegraphics}[2][]{}% Remove \includegraphics functionality
  \renewcommand{\resizebox}[3]{}% Remove \resizebox functionality
}
\setcounter{totalnumber}{10}% Allow up to 10 floats per page
\setcounter{topnumber}{10}% Allow up to 10 floats per page
\setcounter{bottomnumber}{10}% Allow up to 10 floats per page
\renewcommand{\efloatseparator}{\mbox{}}% From the documentation

% Depending on the question, the commands below will prepare the document
% accordingly
\ifthenelse{\boolean{usebiblatex}}{
  %-----------------------------------------------------------------------------
  %  If using biblatex
  %-----------------------------------------------------------------------------
  % Load required packages and refer to correct bibliography
  \usepackage{csquotes}
  \usepackage[
      backend=biber
    , style=apa          % To display Author-year in the text
    , natbib=true        % Necessary to use citep, etc.
    , hyperref=true      % To have clickable links in the references
    , useeditor=false    % Don't add editor to publications
    , sorting=ynt        % Sort by year, name, and title
    , uniquename=false   % Avoid that initials are put
  ]{biblatex}
  \addbibresource{Literature.bib}

  % To include author in link
  \DeclareFieldFormat{citehyperref}{%
  \DeclareFieldAlias{bibhyperref}{noformat}% Avoid nested links
  \bibhyperref{#1}}
  \DeclareFieldFormat{textcitehyperref}{%
  \DeclareFieldAlias{bibhyperref}{noformat}% Avoid nested links
  \bibhyperref{%
    #1%
    \ifbool{cbx:parens}
      {\bibcloseparen\global\boolfalse{cbx:parens}}
      {}}}
  \savebibmacro{cite}
  \savebibmacro{textcite}
  \renewbibmacro*{cite}{%
    \printtext[citehyperref]{%
      \restorebibmacro{cite}%
      \usebibmacro{cite}}}
  \renewbibmacro*{textcite}{%
    \ifboolexpr{
      ( not test {\iffieldundef{prenote}} and
        test {\ifnumequal{\value{citecount}}{1}} )
      or
      ( not test {\iffieldundef{postnote}} and
        test {\ifnumequal{\value{citecount}}{\value{citetotal}}} )
    }
      {\DeclareFieldAlias{textcitehyperref}{noformat}}
      {}%
    \printtext[textcitehyperref]{%
      \restorebibmacro{textcite}%
      \usebibmacro{textcite}}}

  % Remove editor form everything but books
  \AtEveryBibitem{%
   \ifentrytype{book}{}{%
    \clearname{editor}
   }
  }
}{
  %-----------------------------------------------------------------------------
  %  If using bibtex
  %-----------------------------------------------------------------------------
  % Load required packages and select citation style
  \usepackage[round]{natbib}
  \bibliographystyle{apalike}
}

%------------------------------------------------------------------------------
%	Some Styling
%------------------------------------------------------------------------------
% Link colors
% \definecolor{linkcolor}{HTML}{000000}
% \hypersetup{
%   colorlinks = true,
%   linkcolor  = linkcolor,
%   urlcolor   = linkcolor,
%   citecolor  = linkcolor,
% }

@@

% Avoid skip when including text from external .tex file
\newcommand{\inputy}[1]{\input{#1}\unskip}

% For supplementary material
\newcommand{\beginappendix}{%
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}%
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}%
  \setcounter{equation}{0}
  \renewcommand{\theequation}{Equation S\arabic{equation}}%
  \setcounter{section}{0}
  \renewcommand{\thesection}{A.\arabic{section}}%
}

%------------------------------------------------------------------------------
%  ORCID
%------------------------------------------------------------------------------
% Creating some TikZ styles
\tikzset{
  nonterminal/.style = {rectangle
    , minimum size = 6mm
    , very thick
    , draw = black!
  }
}

% Create the ORCID logo
\definecolor{orcidlogocol}{HTML}{A6CE39}
\tikzset{
  orcidlogo/.pic={
    \fill[orcidlogocol]
      svg{M256,128c0,70.7-57.3,128-128,128C57.3,256,0,198.7,0,128C0,57.3,57.3
        ,0,128,0C198.7,0,256,57.3,256,128z};
    \fill[white]
      svg{M86.3,186.2H70.9V79.1h15.4v48.4V186.2z}
      svg{M108.9,79.1h41.6c39.6,0,57,28.3,57,53.6c0,27.5-21.5,53.6-56.8
        ,53.6h-41.8V79.1z M124.3,172.4h24.5c34.9,0,42.9-26.5
        ,42.9-39.7c0-21.5-13.7-39.7-43.7-39.7h-23.7V172.4z}
      svg{M88.7,56.8c0,5.5-4.5,10.1-10.1,10.1c-5.6,0-10.1-4.6-10.1-10.1c0-5.6
      ,4.5-10.1,10.1-10.1C84.2,46.7,88.7,51.3,88.7,56.8z};
  }
}

% Command to create the ORCID
\newcommand\orcid[1]{\href{https://orcid.org/#1}{\mbox{\scalerel*{
\begin{tikzpicture}[yscale=-1,transform shape]
  \pic{orcidlogo};
\end{tikzpicture}
}{|}}}}

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
% \title{Flooding of the Okavango Delta influences Connectivity for Dispersing
% African Wild Dogs}
\title{Dispersal and Connectivity in Increasingly Extreme Climatic Conditions}

% List of Authors
\author{
  David D. Hofmann\textsuperscript{1,2,\S} \orcid{0000-0003-3477-4365} \and
  Dominik M. Behr\textsuperscript{1,2} \orcid{0000-0001-7378-8538} \and
  John W. McNutt\textsuperscript{2} \and
  Arpat Ozgul\textsuperscript{1,2} \orcid{0000-0001-7477-2642} \and
  Gabriele Cozzi\textsuperscript{1,2} \orcid{0000-0002-1744-1940}
}

@@

@@

\textsuperscript{1} Department of Evolutionary Biology and Environmental
Studies, University of Zurich, Winterthurerstrasse 190, 8057 Zurich,
Switzerland.

\textsuperscript{2} Botswana Predator Conservation Program, Wild Entrust,
Private Bag 13, Maun, Botswana.

\textsuperscript{\S} Corresponding author: \href{mailto://david.hofmann2@uzh.ch}{david.hofmann2@uzh.ch}

@@

\textbf{Running Title:} Dispersal in a Changing Climate

@@

\textbf{Keywords:} climate change, conservation, dispersal, flood extremes,
individual-based simulations, landscape connectivity, movement ecology,
step-selection functions

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
While climate change has been shown to impact several life-history traits of
wild-living animal populations, little is known about its effects on dispersal
and connectivity.

Here, we capitalize on the highly variable flooding regime of the Okavango Delta
to investigate impacts of changing environmental conditions on dispersal and
connectivity of the endangered African wild dog (\textit{Lycaon pictus}). Based
on remote sensed flood extents observed over 20 years, we derive two extreme
flood scenarios: a minimum and a maximum flood extent; representative of very
dry and very wet environmental periods. These conditions are akin to those
anticipated under increased climatic variability, as it is expected under
climate change. Using a movement model parametrized with GPS data from
dispersing individuals, we simulate 12,000 individual dispersal trajectories
across the ecosystem under both scenarios and investigate patterns of
connectivity.

Across the entire ecosystem, surface water coverage during maximum flood extent
reduces dispersal success (i.e., the propensity of individuals to disperse
between adjacent subpopulations) by 12\% and increases dispersal durations by
17\%. Locally, however, dispersal success diminishes by as much as 78\%.
Depending on the flood extent, alternative dispersal corridors emerge, some of
which in the immediate vicinity of human-dominated landscapes. Notably, under
maximum flood extent, the number of dispersing trajectories moving into
human-dominated landscapes decreases by 41\% at the Okavango Delta's inflow, but
increases by 126\% at the Delta's distal end. This may drive the amplification
of human-wildlife conflict.

Whilst predicting the impacts of climate change on environmental conditions
on-the-ground remains challenging, our results highlight that environmental
change may have significant consequences for dispersal patterns and
connectivity, and ultimately, population viability. Acknowledging and
anticipating such impacts will be key to effective conservation strategies and
to preserve vital dispersal corridors in light of climate change and other
human-related landscape alterations.

\end{abstract}

@@

% \onehalfspacing
% \tableofcontents
% \doublespacing

@@

Climate change is expected to impact ecosystems across the globe with
far-reaching consequences for the species living therein \citep{Paniw.2019,
Radchuk.2019, IPCC.2022, Ozgul.2023}. By altering environmental conditions,
climate change has been shown to affect animal behavior \citep{Fuller.2016},
resource availability \citep{Durant.2007}, demographic rates of local
populations \citep{Paniw.2021}, range distribution of wild-living animals
\citep{Thomas.2004, Thuiller.2006}, and the potential for human-wildlife
conflict (hereafter referred to as HWC; \citealp{Abrahms.2023}), thus
threatening the viability of most wildlife species. An important life-history
trait known to influence the persistence of wild living populations across large
spatial scales is dispersal \citep{Hanski.1999, Bowler.2005, Kokko.2006}, which
is defined as the movement of individuals away from their natal location to the
site of first reproduction \citep{Clobert.2012}. Despite the crucial role of
dispersal in maintaining population viability, little is known about the impacts
of climate change on long-distance dispersal and landscape connectivity.
Nevertheless, assessing the consequences of climate change for dispersal is
essential to anticipate potential shifts in range distributions and population
dynamics.

Predicting how dispersal and connectivity respond to environmental change
remains challenging \citep{Littlefield.2019}. This is mainly due to insufficient
data on dispersing animals at the appropriate spatial and temporal scales
\citep{Graves.2014, Vasudev.2015}, coupled with limited insights into potential
differences in environmental conditions under climate change
\citep{Scheiter.2009, IPCC.2022}. To address these shortcomings, one approach
has been to combine climate projections and species distribution models with the
aim of predicting future species ranges and investigating the impacts of such
range shifts on structural connectivity \citep{Wasserman.2012, Ashrafzadeh.2019,
Luo.2021}. However, this approach fails to translate predicted atmospheric
conditions into ground-level landscape characteristics (e.g., vegetation cover,
surface-water). An alternative approach, albeit not primarily focusing on
climate change, is to investigate how seasonality affects functional
connectivity (e.g., \citealp{Mui.2017, Osipova.2019, Zeller.2020, Kaszta.2021}).
This is usually achieved using seasonally updated resistance surfaces that
depict the ease or difficulty at which the focal species can traverse a
particular environment in a specific season \citep{Zeller.2012}. Despite its
biological relevance for understanding seasonal variability, this approach
suffers from the short time-span at which processes are investigated and hinders
drawing inferences on the long-term effects of climate change on dispersal and
connectivity. Recently, alternative approaches for investigating landscape
connectivity that combine empirical GPS movement data with a modeling framework
to reconstruct and predict movement trajectories at the level of the individual
have been suggested \citep{Signer.2017, Hofmann.2023, Signer.2023}. Such
predicted trajectories are uncoupled from any temporal constraint and therefore
suitable to investigate the effects of changing landscapes, as observed or
predicted under the influence of climate change, on dispersal and connectivity.
Given that climate change will shift systems towards conditions that currently
represent extremes \citep{Stott.2016, Ummenhofer.2017, IPCC.2022}, we argue that
a focus on extreme, rather than seasonal, variations could serve as a robust way
to learn about the impacts of climate change on patterns of dispersal and
connectivity. Such an approach appears particularly useful in cases where data
on future environmental conditions are difficult to obtain or plagued by
uncertainty \citep{Collins.2012}.

One ecosystem that offers a unique opportunity to study the impacts of extreme
environmental conditions on dispersal and connectivity in a large-scale natural
experiment is the Okavango Delta in Botswana. The Okavango Delta (henceforth
``Delta'') is the world's largest inland delta and characterized by striking
variability in its flood extent within and between years \citep{Gumbricht.2004,
Wolski.2017}. The area covered by the Delta's floodwaters can fluctuate between
3,500 km\textsuperscript{2} during particularly dry and 14,000
km\textsuperscript{2} during particularly wet periods \citep{McCarthy.2003,
Gumbricht.2004}. The region is among the most vulnerable to climate change, as
temperature increases of 4 to 6\degree C above pre-industrial levels are
expected within the 21\textsuperscript{st} century \citep{Engelbrecht.2015,
Akinyemi.2019}; a prediction that goes far beyond the global average
\citep{IPCC.2022}. The rise in ambient temperature is likely to elevate
evapotranspiration rates across the Delta's alluvial fan, yet it is unclear
whether water-losses to the atmosphere will be offset or exceeded by an increase
in precipitation across the Delta's catchment areas in Angola
\citep{Wolski.2008, IPCC.2022}. Despite the importance of the Delta as a driver
of ecosystem functioning \citep{Wolski.2008}, species distribution
\citep{Bonyongo.2005, Bennitt.2014}, and dispersal corridors
\citep{Hofmann.2021, Hofmann.2023}, forecasting its flooding regime under
climate change has proven notoriously difficult. Predictions of the flood extent
under climate change range from ``much drier'' to ``much wetter'' depending on
the employed model and climate scenario \citep{Murray-Hudson.2006, Wolski.2008}.

Within this ecosystem a keystone predator and an umbrella species for
conservation efforts is the African wild dog (\textit{Lycaon pictus}), a large
carnivore that is characterized by a remarkable dispersal ability
\citep{McNutt.1996, Davies-Mostert.2012, Masenga.2016, Cozzi.2020,
Sandoval-Seres.2022}. Under favorable conditions, young individuals that leave
their natal pack can cover up to several hundred kilometers within a few days,
crossing a vast diversity of habitat-conditions (e.g., \citealp{Cozzi.2020}).
While historically observed Euclidean dispersal distances are limited to 5-500
km \citep{Davies-Mostert.2012, Cozzi.2020, Sandoval-Seres.2022}, cumulative
dispersal distances of over 5,000 km have been recorded \citep{Masenga.2016}.
Previous research on dispersing individuals revealed that the Delta's
floodwater represents a major barrier to dispersal \citep{Hofmann.2021,
Hofmann.2023} and it can be hypothesized that wild dog dispersal and
connectivity will be profoundly influenced by future changes in the flood
regime, as expected under the influence of climate change.

Utilizing a previously validated individual-based movement model parametrized
with empirical GPS data collected on dispersing wild dogs, we simulated 12,000
individual dispersal trajectories under two extreme environmental scenarios: one
assuming a maximum flood extent, representing above-average wet climatic
conditions, and one assuming a minimum flood extent, representing acute dry
conditions. Both scenarios reflect possible outcomes under the effect of climate
change. For each scenario, we assessed dispersal success and dispersal durations
of simulated trajectories between distinct regions, mapped the emergence of
alternative dispersal corridors and bottlenecks, and investigated how differing
corridor arrangements influenced the potential for HWC (\Cref{Concept}). We
anticipated lower dispersal success and connectivity, as well as an increased
propensity to disperse outside the main study area during maximum flood.
Furthermore, we anticipated major dispersal corridors to differ between minimum
and maximum flood, thus resulting in alternative hotspots for HWC. Ultimately,
our goal was to highlight that altered climatic conditions and associated
changes in landscape characteristics can substantially affect the spatial
arrangement of movement corridors and, subsequently, the potential for HWC.

\begin{figure}
 \begin{center}
  \includegraphics[width = \textwidth]{Figures/GraphicalAbstract.pdf}
  \caption{Conceptual illustration of the employed analytical framework to
  investigate connectivity for dispersing African wild dogs under a minimum and
  maximum flood extent in the Okavango Delta. (a) Based on a series of 700
  floodmaps, obtained between the years 2000 and 2019, we derived extreme
  scenarios for a minimum and maximum flood, using the minimum 100 and maximum
  100 historically observed flood extremes. These layers were then joined with a
  static set of covariates and fed into a dispersal simulation. (b) The
  dispersal simulation was based on a previously parametrized integrated
  step-selection function for dispersing wild dogs and allowed simulating
  dispersal trajectories in the two extreme scenarios. (c) From simulated
  dispersal trajectories we generated four summary maps, highlighting
  connectivity patterns and areas potentially prone to human-wildlife conflict.}
  \label{Concept}
 \end{center}
\end{figure}

\section{Materials and Methods}

\subsection{Study Area}

To investigate the effects of flood extremes on dispersal and connectivity, we
focused on a \textit{core study area} of 70,000 km\textsuperscript{2} that
comprised the Delta and its surroundings (\Cref{StudyArea}b; purple circle). To
accommodate for the long-distance dispersal events observed in this ecosystem
\citep{Cozzi.2020, Hofmann.2021, Cozzi.2023}, we also considered an
\textit{extended study area} of  300,000 km\textsuperscript{2} spanning from
21\degree 30' S to 17\degree 30' S to and 20\degree 30' E to 26\degree E and
comprising the Delta and parts of Angola, Namibia, Botswana, Zimbabwe, and
Zambia (\Cref{StudyArea}a). Rainfall in this area averages at 450 mm and is
mainly concentrated to the wet season between November and April
\citep{Mendelsohn.2004}. Even though local rains and ground table levels
resulting from floods of previous years may impact annual flood levels
\citep{McCarthy.2006}, the Delta's flood extent is primarily driven by
precipitation patterns in its catchment areas in the Angolan highlands, from
where water is channeled into the Cubango and Cuito rivers and subsequently
discharged into the Okavango Delta \citep{McCarthy.1997, McCarthy.2003}. Since
water only slowly descends from the catchment areas into the Delta and across
its shallow alluvial fan (gradient $\sim$ 1/3,400, \citealp{McCarthy.1997}), the
flood usually reaches its maximum extent long after the local rains have ceased,
during the peak dry season in July or August. Once the floodwater reaches the
Delta's distal ends, it percolates at the Thamalakane and Kunyere Faults, two
natural fault lines at which the water-flow is hindered (\Cref{StudyArea}b).
After reaching its maximum extent, water evaporates or percolates and the flood
steadily retracts over subsequent months. Such inner-annual dynamics are further
amplified or buffered by multi-annual cycles in precipitation patterns across
Angola \citep{Wolski.2012}. Due to the intricate interplay between precipitation
patterns in Angola, local ground-table levels, evapotranspiration, and
anthropogenic water abstractions, predictions of the flood extent under climate
change have proven challenging and range from ``much drier'' to ``much wetter''
depending on employed climate scenarios \citep{Kgathi.2006, Wolski.2008,
Hughes.2011, Wolski.2012, Moses.2018}. In the core study area, the vegetation is
dominated by mopane forest (\textit{Colophospermum mopane}), mixed woodland
acacia-dominated (\textit{Acacia spp.}), and grassland \citep{Mendelsohn.2010}.
Human influence is relatively low and mainly concentrated around small villages
at the western and southern periphery of the Delta. The largest urban center is
Maun, a spread-out settlement at the south-eastern tip of the Delta. Large
portions of the core study area are designated as national parks, game reserves,
or wildlife management areas. Some remaining sections of land are used for
cattle farming and suffer from moderate levels livestock depredation and
associated human-wildlife conflict \citep{Gusset.2009, McNutt.2017}. The core
study area and its extended surroundings are part of the world's largest
transboundary conservation initiative, the Kavango-Zambezi Transfrontier
Conservation Area (KAZA-TFCA, \Cref{StudyArea}a), which attempts to re-establish
connectivity among several core habitats. Previous studies have indicated that
this initiative has high potential for improving connectivity among
subpopulations of various species \citep{Brennan.2020, Lines.2021}, including
African wild dogs \citep{Hofmann.2021}. However, a better understanding of
connectivity within this ecosystem under climate change is lacking.

\begin{figure}
 \begin{center}
  \includegraphics[width = \textwidth]{Figures/StudyArea2.png} \caption{(a)
  Location of the study area, which forms part of the Kavango-Zambezi
  Transfrontier Conservation Area (KAZA-TFCA, red polygon) in Southern Africa.
  (b) The division of the core study area (enclosed by the purple ellipse) into
  sub-areas (1 to 6) was based on the hydrographic structure of the Okavango
  Delta and its tributaries. We simulated dispersal trajectories starting at
  random locations within each of the six source areas (orange circles). Purple
  zones (7 to 14) represent zones that we used to identify if and where
  simulated dispersers left the close surroundings of the Okavango Delta (thus
  called egression zones). These zones were generated using a set of cutlines
  (purple dotted lines) originating from the center of the Delta that dissected
  an elliptical buffer surrounding the Delta into sections of equal size and in
  accordance with cardinal directions. The yellow lines represent GPS
  trajectories of dispersing wild dogs that were observed using GPS collars.
  While not all of these individuals were used to parametrize the dispersal
  model presented by \citet{Hofmann.2023}, they serve to highlight the spatial
  scale at which the species disperses. Map lines delineate study areas and do
  not necessarily depict accepted national boundaries.}
  \label{StudyArea}
 \end{center}
\end{figure}

\subsection{Spatial Habitat Layers}

We represented the physical landscape through which dispersers could move by a
set of spatially referenced habitat layers, each with a resolution of 250 m x
250 m. The set of layers included water-cover, distance-to-water, tree-cover,
shrub/grassland-cover, and a composite human influence layer representing
settlements, roads, and agricultural areas. A detailed description of these
layers is provided by \citet{Hofmann.2021}. However, unlike Hofmann and
colleagues, who used a single, static flood map, we employed composite flood
maps to study connectivity. Specifically, we generated water-cover layers using
MODIS Terra MCD43A4 satellite imagery that we classified into binary water-cover
maps using an algorithm developed by \citet{Wolski.2017} and implemented in
\texttt{R} (\url{https://github.com/DavidDHofmann/floodmapr}). This algorithm
allowed us to generate weekly updated ``flood extent maps'', thus providing
detailed information about the flood at any given point in time. In total, we
generated 700 flood extent maps, covering the years 2000 to 2019. We used these
maps to produce minimum and maximum flood scenarios, representative of dry and
wet climatic conditions, assuming that climate change will shift the system
towards one of these extremes in the long run. To create the baseline minimum
(and respectively, maximum) flood scenario, we averaged the 100 flood extent
maps with the smallest (highest) flood extent and generated a binary layer by
masking all pixels that were inundated in less than 50\% of the maps (see also
\Cref{Concept}). By averaging across 100 flood extent maps, we followed a
conservative approach and mitigated chances of misrepresenting minimum and
maximum flood extent due to inaccuracies or artifacts in single remote sensed
flood extent maps. The final minimum and maximum flood extent maps
(\Cref{FloodExtent}) show that the flood in the two scenarios covers 3,500
km\textsuperscript{2} and 9,500 km\textsuperscript{2}, respectively. Finally, we
combined the two flood extent maps with the set of above-mentioned habitat
layers into two spatial stacks, each representative of one scenario
(\Cref{Concept}). To prevent edge effects during the dispersal simulation, we
followed \citet{Koen.2010} and expanded the stacks to an extent that was 20\%
wider and taller than the original, filling the so created buffer zone (gray
buffer in \Cref{StudyArea}b) with randomized values from the respective layers.
This would allow simulated dispersers to leave and re-enter the main study area
via randomized buffer zones.

\begin{figure}
 \begin{center}
  \includegraphics[width = \textwidth]{Figures/FloodExtent.png}
  \caption{(a) Minimum and (b) maximum flood extent. The two maps were generated
  based on 100 minimum and 100 maximum flood extent maps from a series of 700
  almost weekly updated remote sensed MODIS MCD43A4 satellite images spanning
  the years 2000 to 2019.}
  \label{FloodExtent}
 \end{center}
\end{figure}

\subsection{Source Areas and Egression Zones}

We simulated dispersing wild dogs originating from six distinct \textit{source
areas} located within the main study area (\Cref{StudyArea}b). Defining these
source areas was necessary to enable identification and quantification of the
number of successful dispersal events between different regions of the core
study area across the two flood scenarios. We selected source areas in regions
that remained dry during both scenarios and are known to host viable African
wild dog subpopulations. Nevertheless, these areas are not exhaustive and viable
wild dog populations are known to reside outside of them. The study area
comprises mostly of continuous habitat, yet natural and anthropogenic landscape
features allow the subdivision of the core study area into several sub-regions.
The Delta's hydrography and human settlements at its southern fringes result in
a natural latitudinal split between the eastern (areas 1, 2, and 3) and western
(areas 4 and 5) Delta (\Cref{StudyArea}b). Areas 1 and 2 are separated
longitudinally by the Selinda Spillway, a waterway that connects the Delta with
the Linyanti Swamp, while areas 2 and 3 are separated by the Khwai River, which
inundates the Mababe Depression east of the Delta (\Cref{StudyArea}b). On the
western side of the Delta, areas 4 and 5 are longitudinally separated by the
Thaoga River drainage at Nokaneng (\Cref{StudyArea}b). Finally, the Delta hosts
a central island, known as Chief's Island, which we defined as source area 6
(\Cref{StudyArea}b). The shape of this area was determined by its borders
towards water during the maximum flood scenario, whereas all other source areas
were defined using circular polygons with radii of 20 km at somewhat equal
distances to each other. We deliberately selected equally sized areas to
facilitate comparability between them and to minimize variability in simulated
connectivity due to target-effects. We also avoided source areas that directly
bordered each other, as simulated individuals initiated close to a shared border
would immediately result in connectivity and obscure differences between the two
scenarios. Besides source areas 1 to 6, we also generated \textit{egression
zones} (\Cref{StudyArea}b). These zones enabled us to determine through which
regions simulated individuals left the core study area, as well as to identify
after how many simulated steps they did so. We generated these zones by
overlaying the Delta with an ellipse that we dissected into equally sized
polygons in accordance with cardinal directions (\Cref{StudyArea}b). Overall,
the selection of source areas and egression zones was based on a mixture between
biological and computational considerations.

\subsection{Dispersal Simulation}

We used a previously parameterized and validated dispersal model to simulate
dispersal of African wild dogs. This dispersal model was trained using GPS data
obtained from 16 wild dog coalitions dispersing across northern Botswana
\citep{Hofmann.2023}. These data were collected at a sampling rate of four
hours, with the exception of one GPS location in the afternoon being skipped in
order to accommodate the species tendency to be less active during this period
(details in \citealp{Cozzi.2020, Hofmann.2021}). As such, a total of five GPS
locations were collected per day. The collected data were then fed into an
integrated step-selection function (iSSF, \citealp{Avgar.2016}), where
consecutive GPS locations were converted into \textit{observed steps} (the
straight-line connecting two consecutive GPS recordings; \citealp{Turchin.1998})
and compared to a set of \textit{random steps} using (mixed effects) conditional
logistic regression (\citealp{Fortin.2005, Thurfjell.2014, Muff.2020,
Fieberg.2021}; but see \citealp{Michelot.2024} for alternative approaches).
iSSFs provide less biased estimates of habitat-selection than traditional
resource-selection functions \citep{Forester.2009, Zeller.2016} and bear the
benefit that random steps are drawn from parametric step-length and
turning-angle distributions, such that a model parametrized using iSSFs
resembles a fully mechanistic model from which movement can be simulated
\citep{Signer.2017, Potts.2023, Signer.2023}. In contrast to graph-based
connectivity models (e.g., least-cost path analysis or circuit-theory),
simulations from iSSFs alleviate the need for resistance surfaces, which are
often subjective \citep{Simpkins.2017, Marrec.2020} and overestimate conductance
in difficult to reach habitat \citep{Signer.2017, Hofmann.2023}. The dispersal
model presented by \citet{Hofmann.2023} comprised a habitat-selection function
(describing habitat selection), a movement kernel (describing dispersers'
movement capacities), and potential interactions between the two. According to
the parametrized model, the main characteristics of wild dog dispersal movements
are avoidance of water, avoidance of areas influenced by humans, and a
preference for directional and fast movements. The associated model parameters
(provided in Figure S1) can be used to predict probabilities of a step being
chosen among a set of random steps (details in Supporting Information S1), which
allows to simulate movements in discrete time.

Originating from each of the six source areas (\Cref{StudyArea}b), we simulated
2,000 individual dispersing trajectories, each composed of 2,000 steps. Two
thousand steps corresponded to a dispersal duration of $\sim$ 400 days, which
marks the longest dispersal duration recorded in the study area
\citep{Cozzi.2020, Hofmann.2021}. By simulating 2,000 steps, we generated the
maximum amount of information possible and did not limit connectivity by
defining a dispersal cap. Note that we assumed that individuals would not
disperse for longer than 2,000 steps, but did not model settlement explicitly.
This is because only little is known about how dispersing wild dogs locate
potential mates and a suitable territory to settle. Olfactory cues at shared
scent-marking sites appear to play a crucial role, yet these are extremely
difficult to find and monitor \citep{Apps.2022, Claase.2022}. We simulated 1,000
trajectories under the minimum flood extent, the remaining 1,000 under the
maximum flood extent. Across the six source areas, this resulted in the
simulation of a total of 12,000 individual dispersal trajectories, which
sufficed to achieve convergence in relevant connectivity metrics
\citep{Hofmann.2023}. The simulation procedure was based on the algorithm
described by \citep{Signer.2017} and \citet{Hofmann.2023} and works as follows.
A random location within the source area is defined as the starting point.
Originating from the starting point, a set of 25 random steps is generated by
sampling step lengths ($sl$) from a gamma distribution fitted to observed steps
(shape = 0.37, scale = 6,316) and turning angles ($ta$) from a uniform
distribution ($-\pi, +\pi$). Along each random step, the underlying spatial
covariates are extracted, and relevant movement metrics computed ($log(sl)$ and
$cos(ta)$). \(\beta-\)estimates from the fitted dispersal model are then used to
predict the probability of each step being chosen, given the step's associated
covariates and movement metrics, as well as characteristics of all other
proposed steps. Among the 25 proposed steps, one is chosen at random based on
assigned probabilities. The procedure is then repeated until a total of 2,000
steps is simulated.

\subsection{Derived Metrics}

Based on simulated dispersal trajectories, we quantified connectivity,
identified the emergence of alternative dispersal corridors, and highlighted
areas with elevated potential for HWC. Our approach drew upon the set of
complementary connectivity metrics for individual-based movement models
discussed by \citet{Hofmann.2023}, and was expanded to include a map
highlighting areas with elevated HWC potential. The set of connectivity metrics
comprised \textit{inter-patch connectivity} metrics, summarizing dispersal
success and duration into distinct habitat patches, \textit{heatmaps}, depicting
areas of intense use by dispersers, and \textit{betweenness maps}, highlighting
dispersal corridors and bottlenecks. Finally, the \textit{HWC maps} served to
reveal areas where simulated dispersers moved into the vicinity of
human-dominated landscapes. To illustrate differences between metrics during
maximum and minimum flooding, we computed difference maps for the heatmaps,
betweenness maps, and HWC maps.

\subsubsection{Inter-Patch Connectivity}

To measure inter-patch connectivity, we tallied the number of trajectories
successfully moving from one source-area to another and computed the average
minimum dispersal duration (in number of steps) required to make those
connections. Dispersal between two areas was said to be successful whenever a
trajectory originating from one area intersected with the polygon of another
area. If a trajectory moved through multiple areas, a connection into each of
them from the original source was recorded. We also estimated the number of
individual trajectories that left the core study area and moved into one of the
egression zones. To quantify variability in our estimates, we generated
bootstrap samples by randomly drawing 1,000 simulated trajectories per source
area 1,000 times with replacement and computing 95\%-percentile intervals.

\subsubsection{Intensity of Use}
